---
title: "Example Analysis: Statistical Performance Indicators"
bibliography: refs.bib
---
The SPI data used here originate from the World Bank \[@worldbank2023spi\],
and are accessed via the TidyTuesday dataset curated by Nicola Rennie and the R4DS community \[@tidytuesday2025spi\].

# Background

**The Statistical Performance Indicators (SPI)**, developed by the World Bank, provide a quantitative assessment of how well national statistical systems perform across five dimensions:
	1.	Data use
	2.	Data services
	3.	Data products
	4.	Data sources
	5.	Data infrastructure

The SPI covers nearly all countries worldwide (≈ 99% of the global population) and spans 2004–2023.
It serves as a diagnostic tool to strengthen national data systems.

::: {.callout-note}
Purpose: To monitor and improve national statistical systems’ capacity and performance.
:::
```{r}
#| message: false
#| warning: false

library(tidyverse)
library(scales)

spi <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-12-02/spi_indicators.csv")

glimpse(spi)
```

# Key Questions
	1.	How has a country’s statistical performance changed over time?
	2.	Is statistical performance related to income level or population size?
	3.	Which pillar do countries score lowest in?

::: {.column-margin}
Small differences in SPI between countries may reflect data imprecision rather than true performance gaps.
:::


# Data Wrangling

We clean and simplify the dataset to focus on major regions and income levels.

```{r}
#| warning: false

spi_clean <- spi |>
  filter(!is.na(overall_score)) |>
  mutate(
    income = fct_relevel(income, "Low income", "Lower middle income",
                         "Upper middle income", "High income"),
    region = fct_lump(region, n = 6)
  ) |>
  group_by(region, income, year) |>
  summarise(
    mean_spi = mean(overall_score, na.rm = TRUE),
    .groups = "drop"
  )
```


# Trend in Statistical Performance

```{r}
ggplot(spi_clean, aes(x = year, y = mean_spi, color = income)) +
  geom_line(size = 1) +
  facet_wrap(~ region, scales = "free_y") +
  labs(
    title = "Trends in Statistical Performance by Income and Region",
    subtitle = "World Bank SPI 2004–2023",
    x = "Year",
    y = "Mean Overall SPI Score",
    color = "Income Level",
    caption = "Data source: World Bank SPI, via TidyTuesday 2025-12-02"
  ) +
  theme_minimal()

```



# Relationship with Population Size

We explore whether countries with larger populations have higher or lower SPI.

```{r}
spi_recent <- spi |>
  filter(year == 2023, !is.na(population), !is.na(overall_score))

ggplot(spi_recent, aes(x = population, y = overall_score, color = income)) +
  geom_point(alpha = 0.7) +
  scale_x_log10(
  labels = scales::label_number(
    scale_cut = scales::cut_short_scale()
  )
) +
  labs(
    title = "Statistical Performance vs. Population Size (2023)",
    x = "Population (log scale)",
    y = "Overall SPI Score",
    caption = "Each point = one country"
  ) +
  theme_minimal()
```

# Which Pillar Scores Lowest?

We calculate the average score per pillar across all countries in 2023.

```{r}
spi_pillars <- spi_recent |>
  summarise(
    data_use = mean(data_use_score, na.rm = TRUE),
    data_services = mean(data_services_score, na.rm = TRUE),
    data_products = mean(data_products_score, na.rm = TRUE),
    data_sources = mean(data_sources_score, na.rm = TRUE),
    data_infra = mean(data_infrastructure_score, na.rm = TRUE)
  ) |>
  pivot_longer(everything(), names_to = "pillar", values_to = "score")

ggplot(spi_pillars, aes(x = reorder(pillar, score), y = score, fill = pillar)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Average SPI Pillar Scores (2023)",
    x = "Pillar",
    y = "Average Score",
    caption = "Data: World Bank SPI 2023"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```

# Summary

The SPI data reveal consistent improvement in statistical performance across all regions since 2004, particularly in high-income countries.
However, data infrastructure and data sources pillars remain relatively weak compared with data use or services.
Population size shows no strong correlation with SPI — suggesting that institutional quality matters more than scale.
Future research could link SPI trends with governance or digital infrastructure indicators.

# Function checklist
	•	dplyr: filter(), mutate(), group_by(), summarise(), fct_relevel()
	•	tidyr: pivot_longer()
	•	ggplot2: geom_line(), geom_point(), geom_col(), facet_wrap()

# Reference
